The difference operation between templates of binary cellular automata

### 2015-11-17 article/article.tex
Some changes by Rafael Félix (tks)

\begin{abstract}
The notion of templates has recently been introduced to directly represent families of cellular automata that share certain property. Here we introduce the difference operation between templates, whose applicability is in representing families of binary cellular automata that share some properties but not others. This operation requires another process that generates what we call the \textit{exception templates}, which are also introduced here. The use of both defined operations is illustrated with examples in the space of elementary cellular automata.
\keywords{cellular automata, rule space, templates, difference of templates, exception templates.}
\end{abstract}
%
\section{Introduction}
\label{sec:introducao}
Cellular automata (CAs) are dynamical systems typically discrete in time, space and states.
Through rules of simple local action, CAs can produce behavior of great complexity. \cite{wolfram2002}. The study of classical problems of cellular automata, as the problem of parity \cite{Betel2013} and the density problem \cite{deOliveira2014density}, can help in the understanding of how this complex behavior emerges. %TODO Mudei de local a citação wolfram2002

There are a few ways to seek solutions to these classic problems, the most basic way is test all the rules of a particular family of CAs in order to see if any of these rules solve the problem. However, for large families CAs, which is the most usual situation, this approach is inefficient or impractical.

As research strategy in the larger families, search algorithms are commonly used, in particular those of evolutionary computing, which have proven very effective in density classification problem \cite{wolz2008very}.% "those of" está correto? "which have proven very"?

Another strategy in the search for rules is restricting the search space by rules that have a given property. To represent a subspace with a particular property without the need of enumerate all the rules contained therein can use \textit{templates} idea, as proposed in \cite{deOliveira2014,deOliveira2014b}.

There are a lot of operations that can be applied to templates such as expansion and intersection. In this paper we introduce the operation of \textit{difference between templates} and the operation that generates the \textit{exception templates}. Moreover the application of these operations are presented in some examples.

After the next section presents some basics about cellular automata, Section \ref{sec:templates} presents in more detail what are templates and its major operations. The Section \ref{sec:diferenca_entre_templates_e_templates_de_excecao} introduces the operation of difference between templates and the operation that generates what we call the \textit{exception templates}, as well as their applications. Finally, the Section \ref{sec:consideracoes_finais} makes closing remarks about the news algori

@@

Usually, it is assumed that a cellular automaton cells have $k$ states which are represented by integer values in the range of $[0, k-1]$ The state of a cell is changed by the local function of the automaton (a rule), formed by the state transitions set of a cell, based on its current state and in the states of adjacent cells. To update the values of a cell by local functions, usually a radius of action $r$ is defined and $r$ is the number of adjacent cells to be analyzed in each direction by local functions.

@@

To refer to the rules of a space, it is common to use the number obtained by the outputs decimal representation  of the state transitions, sorting the lexicographically neighborhoods from largest to smallest state; for example, in the elementary space, the number of the rules corresponds to the decimal sequence formed by the 8-bit output, arranged from the neighborhood 111 to 000.%TODO verificar esse trecho. Tinha uma pontuação estranha sobrando

The number of rules in a space is provided by $k^{k^{2r+1}}$, making it easy to realize that any change in the variable $k$ and $r$ generate families with a large amount of rules. A good strategy for handling this problem is the use of static properties, which are properties obtained directly from the set of AC state transitions. The use of static properties can greatly restrict the initial search space. With templates can be represented a set of rules with certain static property.

@@

According to Boccara and Fukś (\cite{boccara2002}), an AC is conservative when each of its local function $f$, applied in every neighborhood $(\alpha_0,\alpha_1, \dots, \alpha_{n-1})$ satisfy the conditions described in Eq. \ref{eq:conservativeCA}.
\begin{equation}
\begin{split}
f(\alpha_0,\alpha_1, \dots,\alpha_{n-1}) = \alpha_0 + (\sum_{i=0}^{n-2}f(0_0,0_1, \dots,0_i,\alpha_1,\alpha_2, \dots,\alpha_{n-1}) \\- f(0_0,0_1, \dots,0_i,\alpha_0,\alpha_1, \dots,\alpha_{n-i-1}))
\label{eq:conservativeCA}
\end{split}
\end{equation}

@@

Given the transition table of an AC, there are three changes that can employed resulting in a dynamically equivalent behavior to ACs: \textit{reflection}, \textit{combination} and \textit{composition}. Reflection is the transformation obtained reflecting the bits of the neighborhoods transitions table. The conjugation is obtained by reversing all the states of the cells of the transitions table. Finally, the composition is obtained making reflection and conjugation, regardless of the order.

@@

The internal symmetry is defined by the number of state transitions which remain identical after applied a specific transformation. For example, the rule 60 from the ACEs has a value of internal symmetry by reflection equal to 4, because it shares the state transitions $((1,1,1),0)$, $((1,0,1),1)$, $((0,1,0),1)$ e $ ((0,0,0),0)$ with the resulting rule of its transformation by reflection, the rule 102.

\subsection{Color blind cellular automata}
It is considered that an AC has the color blind property if it is invariant to the application of permutations in the states of transition table \cite{salo2013color}. Naturally, there is generally $k!$ permutations, but only one in the binary case, which can be described as $\{0 \to 1 ,\, 1 \to 0\}$.

@@

\section{Templates}
\label{sec:templates}
Template is a generalization of the state transition tables of CAs through variables. A single template may represent a rule set that may present particular property. The templates were created and implemented by De Oliveira and Verardo (2014) and are currently available at the library \textit{open source CATemplates} \cite{CATemplates} on GitHub.

@@

The difference operation takes two templates as a parameter, which we will call $T_{minuend}$ and $T_{subtrahend}$. This operation results in a set of templates that represents all the rules represented by the template $T_{minuend}$ hat are not represented by the $T_{subtrahend}$.

The difference operation is a process with several steps. The first is to do the intersection between the two templates passed as parameter, resulting in a template $T_i$. If there is no intersection between the two original template, the result of the difference is the actual $T_{minuend}$. If there intersection, the intersection of template $T_i$ is matched to the template $T_{minuend}$, thereby generating logical combinations of equations. Then the algorithm removes the tautological equations and applies a negation operation in the equation, which in the binary case is to just make the permutations $\rho = (0 \to 1, 1 \to 0)$ to the result of the equations. Then, the logical operator is exchanged from $\wedge$ to $\vee$ and the system generated by this process is solved, thereby generating a set of replacement sets that should be applied to the template $T_{minuend}$. If the replacement set is empty or invalid (ie, reference non existing rules in the space, as exemplified below), all the rules pertaining to $T_{minuend}$ also belong to the $T_{subtrahend}$ and, therefore, the algorithm returns an empty set.

@@

Finally, apply a negation operation on the equations that, in the binary case, corresponding to the permutation $\rho $ or, equivalently, the implementation of the function $f(x) = 1 - (x)$. Eq. \ref{eq:differenceP4} represents the logical combination of equations resulting from these operations.
\begin{equation}
1 - x_0 = 1 - 1         \vee 
1 - x_3 = 1 - (1 - x_1 - x_2) \vee
x_3   = 1 - (x_1 + x_2)   \vee 
x_0   = 1 - 0       
\label{eq:differenceP4}
\end{equation}

@@

In many cases only the steps described so far are sufficient to make the difference between both templates. But there are cases in which the template $T_{subtrahend}$, corresponding to $T_{con}$ in the given example, the variables has substitutions that lead to invalid rules. This situation occurs, for example, when expanding the template $(1, 1 - x_1, 1 - x_2, 1 - x_1 - x_2, x_1 + x_2, x_2, x_1, 0)$ by assigning the value $1$ to the variables $x_1$ and $x_2$. By doing this is obtained in position $3$ (from right to left) the value of $2$, which is outside the entire range $[0, k-1]$, corresponding therefore to a rule that is not part of the treated space.

@@

Therefore every rule represented  by the template $T_{minuend}$ -- $T_{inv}$ in the given example -- and which is also represented by the exception template from the the intersection of $T_{minuend}$ with $T_{subtrahend}$ -- $T_i$ in this example - - should be represented by at least one of the templates resulting from the difference operation. For this, the algorithm that finds the difference between templates considers all exception templates found, intersects them with the $T_{minuend}$ and adds them to the set of templates obtained by the first steps of the difference operation. Thus, in the example used so far, the set of difference templates resultant is represented by Eq. \ref{eq:differenceR}:\begin{equation}
\begin{split}
\{(0, 1 - x_1, 1 - x_2, 1 - x_3, x_3, x_2, x_1, 1), \\
(1 - x_0, 1 - x_1, 1 - x_2, x_1 + x_2, 1 - x_1 - x_2, x_2, x_1, x_0), \\
(1 - x_0, 0, 0, 1 - x_3, x_3, 1, 1, x_0)\}
\label{eq:differenceR}
\end{split}
\end{equation}

@@

\section{Final Remarks}
\label{sec:consideracoes_finais}
This article describes the templates of cellular automata and presents for the first time the difference operation between templates and the operation that finds exception templates. Both operations could prove valuable in many situations involving to searching a rule in a large space, particularly in the classic problems of parity and density classification.

It is also shown how it is possible to obtain non-trivial answers to questions about searches for rules with certain static properties by means of templates, without relying on the use of search algorithms or enumerative reviews in a whole family of CAs.

It is noteworthy that the difference operation can generate a large number of templates, which can, in specific cases, not effectively reduce the search space. But note that the approach is relevant in general terms, and so more effective the greater the complexity of the attributes specified in the rules, ie, the greater the amount of property that the rules space in question should either not present. Though it was possible to specify the desired properties by means of templates, by the previous work \cite{deOliveira2014,deOliveira2014b}, with the present work will be possible also specifying unwanted properties.

For now the difference operation between templates has been implemented just to the families of ACs with $k=2$, but its generalization to higher values of $k$ is underway.

\section*{Acknowledgements}
\label{sec:agrdecimentos}
The authors thank Mackenzie Research Fund (MACKPESQUISA), the Foundation for State of São Paulo (FAPESP), and federal agencies CAPES and CNPq for different forms of support received during the development of this work.

>>>>

\begin{abstract}
The templates consist in a notation that has recently been introduced to represents families of cellular automata that share some property. This paper introduces the operation of difference between templates. This operation is applicable to representation of binary cellular automata families that share some properties but not others. This papers also introduce a process called \textit{exception templates} that is required to this operation. Experimental results and examples of both techniques defined above are illustrated with examples in the space of elementary cellular automata.
\keywords{cellular automata, rule space, templates, difference between templates, exception templates.}
\end{abstract}
%
\section{Introduction}
\label{sec:introducao}
Cellular automata (CAs) are dynamical systems typically discrete in time, space and states.
CAs can produce behavior of great complexity based on simple rules of local action \cite{wolfram2002}. The study of classical problems of cellular automata, as the problem of parity \cite{Betel2013} and the density problem \cite{deOliveira2014density}, can help in the understanding of how this complex behavior emerges.

The most basic approach to find a suitable solution to classic problems consist in test each of rules of a particular family of CAs in order to see if any of these rules are able to solve the problem. However, this approach shows to be inefficient or impractical for a large number of families in CAs, which is the most usual panorama in literature.

Evolutionary computing is constantly used to deal with larger families. This type of algorithm have proven itself very effective in density classification problem \cite{wolz2008very}.

Another strategy to find suitable rules is restricting the search space by rules that have a given property. To represent a subspace with a particular property without the need of enumerate all the rules contained therein can use \textit{templates} idea, as proposed in \cite{deOliveira2014,deOliveira2014b}.

There are a lot of operations that can be applied to templates such as expansion and intersection. In this paper we introduce the operation named \textit{difference between templates} and the operation that generates the \textit{exception templates}. Moreover, application of these operations are presented in this paper with practical examples.

This paper is organized in sections. The next section, \ref{sec:automatos_celulares}, presents some basic concepts about cellular automata. Following, Section \ref{sec:templates} presents more details about templates and its major operations. The Section \ref{sec:diferenca_entre_templates_e_templates_de_excecao} introduces the operation of difference between templates and the operation that generates what we call the \textit{exception templates}, as well as their applications. Finally, the Section \ref{sec:consideracoes_finais} makes closing remarks about the news algorithm.

@@

Conventionally, it is assumed that a cellular automaton cells have $k$ states which are represented by integer values in the range of $[0, k-1]$ The state of a cell is changed by the local function of the automaton (a rule), formed by the state transitions set of a cell, based on its current state and in the states of adjacent cells. To update the values of a cell by local functions, usually a radius of action $r$ is defined and $r$ is the number of adjacent cells to be analyzed in each direction by local functions.

@@

To refer to the rules of a space, it is common to use the number obtained by the outputs decimal representation  of the state transitions, sorting the lexicographically neighborhoods from largest to smallest state; for instance, in the elementary space, the number of the rules corresponds to the decimal sequence formed by the 8-bit output, arranged from the neighborhood 111 to 000.

The number of rules in a space is provided by $k^{k^{2r+1}}$, making it easy to realize that any change in the variable $k$ and $r$ generate families with a large amount of rules. An approach to handle this problem is to use static properties. The static properties consists of properties obtained directly from the set of CA state transitions. The use of static properties can highly restrict the initial search space. With templates can be represented a set of rules with certain static property.

@@

According to Boccara and Fukś (\cite{boccara2002}), a CA is conservative when each of its local function $f$, applied in every neighborhood $(\alpha_0,\alpha_1, \dots, \alpha_{n-1})$ satisfy the conditions described in Eq. \ref{eq:conservativeCA}.
\begin{equation}
\begin{split}
f(\alpha_0,\alpha_1, \dots,\alpha_{n-1}) = \alpha_0 + (\sum_{i=0}^{n-2}f(0_0,0_1, \dots,0_i,\alpha_1,\alpha_2, \dots,\alpha_{n-1}) \\- f(0_0,0_1, \dots,0_i,\alpha_0,\alpha_1, \dots,\alpha_{n-i-1}))
\label{eq:conservativeCA}
\end{split}
\end{equation}

@@

Given a transitions table of an CA, there are three changes that can be applied and results in a dynamically equivalent behavior to CAs: \textit{reflection}, \textit{combination} and \textit{composition}. Reflection is the transformation obtained reflecting the bits of the neighborhoods transitions table. The conjugation is obtained by reversing all the states of the cells of the transitions table. Finally, the composition is obtained making reflection and conjugation, regardless of the order.

@@

The internal symmetry is defined by the number of state transitions which remain identical after applied a specific transformation. For example, the rule 60 from the elementary CAs has a value of internal symmetry by reflection equal to 4, because it shares the state transitions $((1,1,1),0)$, $((1,0,1),1)$, $((0,1,0),1)$ e $ ((0,0,0),0)$ with the resulting rule of its transformation by reflection, the rule 102.

\subsection{Color blind cellular automata}
A CA is considered color blind when it is invariant to the application of permutations in the states of transition table \cite{salo2013color}. Naturally, there is generally $k!$ permutations, but only one in the binary case, which can be described as $\{0 \to 1 ,\, 1 \to 0\}$.

@@

\section{Templates}
\label{sec:templates}
Template is a generalization of the state transition tables of CAs through variables. A template can represents a rule set, which in turn can presents particular property. The templates were created and implemented by De Oliveira and Verardo (2014) and are currently available at the library \textit{open source CATemplates} \cite{CATemplates} on GitHub.

@@

The difference operation has two templates as an input parameter, which we will call $T_{minuend}$ and $T_{subtrahend}$. This operation output will result in a set of templates that represents all the rules represented by the template $T_{minuend}$ and are not represented by the $T_{subtrahend}$.

The difference operation is a process with several steps. The first consists in a process of intersection between the two selected templates set as parameters, resulting in a template $T_i$. If there is no intersection between the two original template, the result of the difference is the actual $T_{minuend}$. If there intersection, the intersection of template $T_i$ is matched to the template $T_{minuend}$, thereby generating logical combinations of equations. Then the algorithm removes the tautological equations and applies a negation operation in the equation, which in the binary case is to just make the permutations $\rho = (0 \to 1, 1 \to 0)$ to the result of the equations. Then, the logical operator is exchanged from $\wedge$ to $\vee$ and the system generated by this process is solved, thereby generating a set of replacement sets that should be applied to the template $T_{minuend}$. If the replacement set is empty or invalid (ie, reference non existing rules in the space, as exemplified below), all the rules pertaining to $T_{minuend}$ also belong to the $T_{subtrahend}$ and, therefore, the algorithm returns an empty set.

@@

The last step, apply a negation operation on the equations that, in the binary case, corresponding to the permutation $\rho $ or, equivalently, the implementation of the function $f(x) = 1 - (x)$. Eq. \ref{eq:differenceP4} represents the logical combination of equations resulting from these operations.
\begin{equation}
1 - x_0 = 1 - 1         \vee 
1 - x_3 = 1 - (1 - x_1 - x_2) \vee
x_3   = 1 - (x_1 + x_2)   \vee 
x_0   = 1 - 0       
\label{eq:differenceP4}
\end{equation}

@@

In many cases, only the steps described so far are sufficient to make the difference between both templates. But there are cases in which the template $T_{subtrahend}$, corresponding to $T_{con}$ in the given example, the variables has substitutions that lead to invalid rules. This situation occurs, for example, when expanding the template $(1, 1 - x_1, 1 - x_2, 1 - x_1 - x_2, x_1 + x_2, x_2, x_1, 0)$ by assigning the value $1$ to the variables $x_1$ and $x_2$. By doing this is obtained in position $3$ (from right to left) the value of $2$, which is outside the entire range $[0, k-1]$, corresponding therefore to a rule that is not part of the treated space.

@@

Therefore every rule represented  by the template $T_{minuend}$ -- $T_{inv}$ in the given example -- and which is also represented by the exception template from the the intersection of $T_{minuend}$ with $T_{subtrahend}$ -- $T_i$ in this example - - should be represented by at least one of the templates resulting from the difference operation. For this, the algorithm that finds the difference between templates considers all exception templates found, intersects them with the $T_{minuend}$ and adds them to the set of templates obtained by the first steps of the difference operation. Thus, for the example presented above, the set of difference templates resultant is represented by Eq. \ref{eq:differenceR}:\begin{equation}
\begin{split}
\{(0, 1 - x_1, 1 - x_2, 1 - x_3, x_3, x_2, x_1, 1), \\
(1 - x_0, 1 - x_1, 1 - x_2, x_1 + x_2, 1 - x_1 - x_2, x_2, x_1, x_0), \\
(1 - x_0, 0, 0, 1 - x_3, x_3, 1, 1, x_0)\}
\label{eq:differenceR}
\end{split}
\end{equation}

@@

\section{Final Remarks}
\label{sec:consideracoes_finais}
This paper describes the templates of cellular automata, 
introduces the difference operation and show an approach to find exception templates. Both operations could prove valuable in many situations involving efforts to find rules in a large space, particularly in the classic problems of parity and density classification.

Also, it is shown the possibility of obtain non-trivial answers to questions about searches for rules with certain static properties by means of templates, without relying on the use of search algorithms or enumerative reviews in a whole family of CAs.

It is noteworthy that the difference operation can generate a large number of templates, which can, in specific cases, not effectively reduce the search space. But the approach is relevant in general terms, and so more effective the greater the complexity of the attributes specified in the rules, ie, the greater the amount of property that the rules space in question should either not present. Though it was possible to specify the desired properties by means of templates, by the previous work \cite{deOliveira2014,deOliveira2014b}, with the present work will be possible also specifying unwanted properties.

For now the difference operation between templates has been implemented just to the families of CAs with $k=2$, but its generalization to higher values of $k$ is underway.

\section*{Acknowledgements}
\label{sec:agrdecimentos}
The authors thank Mackenzie Research Fund (MACKPESQUISA), the FAPESP, and federal agencies CAPES and CNPq for different forms of support received during the development of this work.
